From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-07-30 WOLFHPC15/article/introduction.tex
add a fix?
intro

Programming complex scientific applications on high performance heterogenous architectures, such as clusters of nodes containing multicores and accelerators, requires for the scientists to be an expert in his own domain, but also to be an expert in low level parallel and high performance programming. It can takes months to years to a scientist to learn HPC programming, as it is not his main activity. And while learning HPC programming, hardware continue to evolve, becoming more powerful but usually also more  complex and difficult to use efficiently.

Domain Specific Languages (DSLs) are a promising solution to hide intricate details of HPC programming from non experts, while producing automatically an optimized and efficient final code. Thus, it seems that a clue for simplicity and efficiency on modern architectures is to specialize a language to a given domain, from which one or more efficient parallelizations and optimizations can be extracted.
However, if the specific nature of a DSL simplifies HPC programming for scientists, its implementation cost (language, compiler, and runtime) is difficult to amortize. Actually, programming difficulties, as well as non-maintainability and non-portability seems to be deported from the scientist to DSL designers. Moreover, one can consider that end users need as much DSLs as domains, which keeps the amount of work to write and maintain up-to-date DSLs out of hands.

For this reason, solutions to ease the design and implementation of DSLs have recently appear~\cite{Fernandez:2014:DFL:2691166.2691168}. The main idea of those solutions is to propose a single DSL to write new DSLs. In this paper, we study a first contribution to another approach to bring maintainability and code-reuse in DSLs. This approach proposes a transformation from a DSL to a component-based runtime. TO DEV BENEFITS OF COMPONENTS.

To evaluate such an approach, a proper new DSL for \emph{multi-stencil} programs, and its compiler are presented in this paper. Many solutions to automatically optimize and parallelize partial differential equations (PDEs) solvers have been proposed in the litterature, as specific libraries~\cite{petsc-efficient,Trilinos-Overview,CPE:CPE3494}, and DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focuss on the parallelization and the optimization of a single numerical computation, also called a stencil.
However, a real case numerical simulation is most of the time not composed of a single stencil computation, but of a set of stencil computations and a set of additionnal auxiliary local computations. The DSL proposed in this paper, called MSL, is a DSL to generate a parallel component-based structure of an overall numerical simulation, that we call a \emph{multi-stencil programs}. As it will be detailed in the related work, this work is complementary to the optimization and parallelization of a single stencil computation.

The rest of this paper is organized as follows. First concepts of \emph{stencil} and \emph{multi-stencil porgams} are formalized in Section~\ref{sect:concept}. Section~\ref{sect:mscac} gives an overview of the solution presented in this paper while Section~\ref{sect:msmsc} details the MSL language and its compiler. Section~\ref{sect:component} introduces component models and then it describes the proposed component-based runtime and the automatic generation of its parallel compuration part.
%The end of Section~\ref{sect:component} also discusses on interest of this new approach.
Section~\ref{sect:eval} focuses on a real-case simulation, solving the Shallow-water equations, and analyzes some performance results. Sections~\ref{sect:related} states on related work, and finally Section~\ref{sect:conclusion} concludes and gives some perspectives.

>>>>

Programming complex scientific applications on high performance heterogeneous architectures, such as clusters of multicores nodes containing accelerators, requires for the scientists to be an expert in his own domain, but also in low level parallel and high performance programming. It can take months to years for a scientist to learn HPC programming, as it is not his main activity. And while learning HPC programming, hardware continue to evolve, becoming more powerful but usually also more  complex and difficult to use efficiently.

Domain Specific Languages (DSLs) are a promising solution to hide intricate details of HPC programming from non experts.
By being specific to a well-defined set of problems, these languages enable the generation of well parallelized and optimized code for various architectures while requiring only a limited amount of information from the developer; part of the knowledge is directly embedded in the compiler.
A critical question however concerns the specificity of DSLs.
The less specific the DSL is, the more informations its users must provide and the less automatic optimizations can be embedded in the compiler.
On the other side of the scope, the more specific the DSL is, the less different code can use it and the more difficult it is to amortize the implementation cost of the DSL (language, compiler, and runtime).
Programming difficulties, as well as non-maintainability and non-portability often seem to be simply deported from the scientist to DSL designers.
% Moreover, one can consider that end users need as much DSLs as domains, which keeps the amount of work to write and maintain up-to-date DSLs out of hands.

For this reason, solutions to ease the design and implementation of DSLs have recently appeared~\cite{Fernandez:2014:DFL:2691166.2691168}. The main idea of those solutions is to propose a single DSL to write new DSLs. In this paper, we study a first contribution to another approach to improve maintainability and code-reuse in DSLs. This approach proposes a transformation from a DSL to a component-based runtime.
To evaluate this approach, a proper new DSL for \emph{multi-stencil} programs, and its compiler are presented in this paper.
\fix{cp2cp+hc: to dev benefits of components.}

Many solutions to automatically optimize and parallelize partial differential equations (PDEs) solvers have been proposed in the literature, either as specific libraries~\cite{petsc-efficient,Trilinos-Overview,CPE:CPE3494} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil.
However, a real case numerical simulation is most of the time not composed of a single stencil computation, but of a set of stencil computations and a set of additional local auxiliary computations. The DSL proposed in this paper, called MSL, is a DSL to generate a parallel component-based structure of an overall numerical simulation, that we call a \emph{multi-stencil program}. As it will be detailed in the related work, this is complementary to the optimization and parallelization of a single stencil computation.
\fix{JB: j'utiliserais bien le terme "kernel" ou "stencil kernel" pour ce que tu apelles "stencil" et "stencil" tout court pour ce que tu apelles "stencil shape"}

The rest of this paper is organized as follows. First, the concepts of \emph{stencil} and \emph{multi-stencil programs} are formalized in Section~\ref{sect:concept}. Section~\ref{sect:mscac} gives an overview of the solution presented in this paper while Section~\ref{sect:msmsc} details the MSL language and its compiler. Section~\ref{sect:component} introduces component models and describes the proposed component-based runtime and the automatic generation of its parallel computation part.
%The end of Section~\ref{sect:component} also discusses on interest of this new approach.
Section~\ref{sect:eval} focuses on a real-case simulation, solving the Shallow-water equations, and analyzes some performance results. Sections~\ref{sect:related} states on related work, and finally Section~\ref{sect:conclusion} concludes and gives some perspectives.
